---
title: "Generate Figures for Toxo RNA-seq Data"
output: html_notebook
---

Zoë Steier

Generate figures for bulk RNA-seq data from mice infected with Toxoplasma gondii. T cells were isolated and sorted for three cell types: A (memory), B (intermediate), and C (effector). There are 10 samples from GRA6 mice (polyclonal TCR).

# Load packages

```{r Load required packages}

library(tidyverse)
library(DESeq2)
library(tximport)
library(readxl)
library(pheatmap)
library(RColorBrewer)

# library(devtools)
# devtools::install_github("saeyslab/triwise")
# install.packages("dplyr") # to get if_any
library(triwise)
```

# Load data

```{r Load RSEM RNA-seq data}

# Load RSEM Count Data
### Note: future users should obtain this data from GEO: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE187545

txi.rsem <- readRDS("/data/yosef2/users/zsteier/Toxo/analysis/GEO_data/txi_rsem_transcript_counts.rds")
```

```{r Make sample metadata}

samples <- colnames(txi.rsem$counts)
GRA6_meta <- tibble(sample = samples) %>%
    mutate(mouse = str_sub(sample, 1,6), cell_type = str_sub(sample,7, 7))

GRA6_meta[["cell_type"]] <- str_replace(GRA6_meta[["cell_type"]], "A", "memory")
GRA6_meta[["cell_type"]] <- str_replace(GRA6_meta[["cell_type"]], "B", "intermediate")
GRA6_meta[["cell_type"]] <- str_replace(GRA6_meta[["cell_type"]], "C", "effector")

meta_df <- data.frame(GRA6_meta) # make dataframe of metadata
row.names(meta_df) <- GRA6_meta$sample
```

```{r Load DE results}

# GRA6 DE results from all pairwise comparisons
GRA6_DE <- read_csv("../DE/DE_results/DESeq2_Toxo_GRA6.csv")
GRA6_DE
```

```{r Set up DESeq2 dataset}

dedsRNASeq_GRA6 <- DESeqDataSetFromTximport(txi.rsem,
                                 colData = meta_df,
                                 design = ~ cell_type
                                 )

# Filter based on minimum reads
keep <- rowSums(counts(dedsRNASeq_GRA6)) >= 10 # Keep genes that have at least 10 reads total
dedsRNASeq_GRA6_filt <- dedsRNASeq_GRA6[keep, ]
```

# Plot figures
## PCA

```{r PCA plotting function}

PlotPCA_celltype <- function(dds, title){
  
  vsd <- vst(dds, blind=FALSE) #dds = dedsRNASeq
  pcaData <- plotPCA(vsd, intgroup=c("cell_type", "mouse"), returnData=TRUE)
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  print(paste0("PC1: ",percentVar[1],"% variance"))
  print(paste0("PC2: ",percentVar[2],"% variance"))
  
  ggplot(pcaData, aes(PC1, PC2, shape = cell_type, color = cell_type)) +
    geom_point(size=3) +
    scale_color_manual(values = c(effector = "red", intermediate = "blue", memory = "black")) + # set colors
    scale_shape_manual(values = c(effector = 17, intermediate = 15, memory = 16)) + # triangle, square, circle
    xlab("PC1") +
    ylab("PC2") +
    coord_fixed() +
    labs(title = title, color = "Cell type", shape = "Cell type", xlabel = "PC1", ylabel = "PC2") +
    theme_classic()
}
```

```{r Plot PCA}

# Plot PCA
PlotPCA_celltype(dedsRNASeq_GRA6_filt, "GRA6")
ggsave("figures/PCA_GRA6.png", width = 5, height = 5)
```

## Heatmap of all DE genes

```{r Find DE genes}

# DE if padj < 0.05 and lfc > 1
de_mem_int <- GRA6_DE %>%
  filter((memory_vs_intermediate_padj < .05) & (abs(memory_vs_intermediate_log2FoldChange) > 1)) %>%
  dplyr::select(Gene) %>%
  pull()

de_mem_eff <- GRA6_DE %>%
  filter((memory_vs_effector_padj < .05) & (abs(memory_vs_effector_log2FoldChange) > 1)) %>%
  dplyr::select(Gene) %>%
  pull()
         
de_int_eff <- GRA6_DE %>%
  filter((intermediate_vs_effector_padj < .05) & (abs(intermediate_vs_effector_log2FoldChange) > 1)) %>%
  dplyr::select(Gene) %>%
  pull()

all_de_gra6 <- unique(c(de_mem_int, de_mem_eff, de_int_eff))

length(all_de_gra6)
length(de_mem_int)
length(de_mem_eff)
length(de_int_eff)
```

```{r Make three annotation columns for DE}

# Find up/down DE genes in each test
de_annotation_tibble <- GRA6_DE %>%
  filter(Gene %in% all_de_gra6) %>%
  mutate(DE_memory_vs_intermediate = case_when( # add annotation 1
    (memory_vs_intermediate_padj < .05) & (memory_vs_intermediate_log2FoldChange > 1) ~ "DE_memory",
    (memory_vs_intermediate_padj < .05) & (memory_vs_intermediate_log2FoldChange < -1) ~ "DE_intermediate",
    TRUE ~ "Not_DE")) %>%
  mutate(DE_memory_vs_effector = case_when( # add annotation 2
    (memory_vs_effector_padj < .05) & (memory_vs_effector_log2FoldChange > 1) ~ "DE_memory",
    (memory_vs_effector_padj < .05) & (memory_vs_effector_log2FoldChange < -1) ~ "DE_effector",
    TRUE ~ "Not_DE")) %>%
  mutate(DE_intermediate_vs_effector = case_when( # add annotation 3
    (intermediate_vs_effector_padj < .05) & (intermediate_vs_effector_log2FoldChange > 1) ~ "DE_intermediate",
    (intermediate_vs_effector_padj < .05) & (intermediate_vs_effector_log2FoldChange < -1) ~ "DE_effector",
    TRUE ~ "Not_DE")) %>%
  select(Gene,  DE_memory_vs_intermediate, DE_intermediate_vs_effector, DE_memory_vs_effector)

row_annotations_df <- data.frame(de_annotation_tibble[2:4])
row.names(row_annotations_df) <- de_annotation_tibble$Gene
```

```{r Summarize GRA6 DE genes to plot}

vsd_GRA6 <- vst(dedsRNASeq_GRA6_filt, blind=FALSE) # Take vst normalization

# Combine all data into one tibble for plotting
heatdata_t <- as_tibble(rownames_to_column(as.data.frame(t(assay(vsd_GRA6))), var = "Sample_name")) # samples x genes

# Select genes of interest here to make the size of the dataframe much smaller
heatdata_t <- heatdata_t %>%
  select(Sample_name, all_of(all_de_gra6))

heatdata_tnew <- heatdata_t %>%
  left_join(GRA6_meta %>% select(sample, cell_type), by = c("Sample_name" = "sample")) %>%
  dplyr::select(-c(Sample_name)) %>%
  select(Cell_type = cell_type, everything())

# Extract annotations, now in the correct order
col_annotations <- data.frame(heatdata_tnew %>% dplyr::select(Cell_type))
# Give annotations row names that match column names of heatdata
row.names(col_annotations) <- heatdata_t$Sample_name
annotation_colors <- list("Cell_type" = c(memory = "black", intermediate = "blue", effector = "red"),
                          "DE_memory_vs_effector" = c(Not_DE = "gray", DE_memory = "black", DE_effector = "red"),
                          "DE_intermediate_vs_effector" = c(Not_DE = "gray", DE_intermediate = "blue", DE_effector = "red"),
                          "DE_memory_vs_intermediate" = c(Not_DE = "gray", DE_memory = "black", DE_intermediate = "blue")
                          )

# Transpose gene expression and rename samples
heatdata <- t(heatdata_tnew %>% select(-c(Cell_type)))
colnames(heatdata) <- heatdata_t$Sample_name
```

```{r Plot heatmap of DE genes}

de_heatmap <- pheatmap(heatdata,
         cluster_rows = TRUE,
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = TRUE,
         annotation_col = col_annotations,
         angle_col = 90,
         annotation_colors = annotation_colors,
         annotation_row = row_annotations_df,
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         scale = "row", # scale colors per row
         cellheight = .08, 
         cellwidth = 15, 
         treeheight_col = 10,
         treeheight_row = 50 
  )
```


```{r Arrange columns}

# Change the order to memory > intermediate > effector by reversing the order above
de_order_cols <- rev(colnames(heatdata[,de_heatmap$tree_col[["order"]]]))
col_annotations_arranged_de <- data.frame(Cell_type = factor(col_annotations[de_order_cols,], levels = c("memory", "intermediate", "effector")))
row.names(col_annotations_arranged_de) <- de_order_cols
```

```{r Arrange rows by annotation}
row_arrange_heatdata <- as_tibble(heatdata, rownames = "Gene") %>%
  left_join(de_annotation_tibble, by = "Gene") %>%
  arrange(DE_memory_vs_effector, DE_intermediate_vs_effector, DE_memory_vs_intermediate) %>%
  select(Gene) %>%
  pull()

# Save
pheatmap(heatdata[row_arrange_heatdata, de_order_cols],
         cluster_rows = FALSE,
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = col_annotations_arranged_de,
         angle_col = 90,
         annotation_colors = annotation_colors,
         annotation_row = row_annotations_df,
         annotation_names_col = TRUE,
         annotation_names_row = TRUE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100),
         scale = "row", # scale colors per row
         cellheight = .1, 
         cellwidth = 15, 
         filename = str_c("figures/heatmap_GRA6_DEgenes.png")
  )

# View
pheatmap(heatdata[row_arrange_heatdata, de_order_cols],
         cluster_rows = FALSE,
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = col_annotations_arranged_de,
         angle_col = 90,
         annotation_colors = annotation_colors,
         annotation_row = row_annotations_df,
         annotation_names_col = TRUE,
         annotation_names_row = TRUE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100),
         scale = "row",
         cellheight = .05, 
         cellwidth = 15 
  )
```


## Heatmap of selected genes

```{r Load gene list}

selected_genes <- read_csv("selected_genes.csv") %>%
  mutate(Gene = str_to_title(Gene))
```

```{r Select genes}

# Select genes to plot
selected_genes_Main1e <- selected_genes %>%
  filter(Figure == "Main1e")

selected_genes_Main1g <- selected_genes %>%
  filter(Figure == "Main1g")

selected_genes_Supp1e <- selected_genes %>%
  filter(Figure == "Supp1e")
```

```{r Select expression data to plot}

# Get expression data
heatdata_t <- as_tibble(rownames_to_column(as.data.frame(t(assay(vsd_GRA6))), var = "Sample_name")) # samples x genes

# Select genes of interest here to make the size of the dataframe much smaller
heatdata_t_Main1e <- heatdata_t %>%
  dplyr::select(Sample_name, all_of(selected_genes_Main1e$Gene[selected_genes_Main1e$Gene %in% colnames(heatdata_t)])) %>%
  left_join(GRA6_meta %>% dplyr::select(sample, cell_type), by = c("Sample_name" = "sample")) %>%
  dplyr::select(-c(Sample_name)) %>%
  dplyr::select(Cell_type = cell_type, everything())

heatdata_t_Main1g <- heatdata_t %>%
  dplyr::select(Sample_name, all_of(selected_genes_Main1g$Gene[selected_genes_Main1g$Gene %in% colnames(heatdata_t)])) %>%
  left_join(GRA6_meta %>% dplyr::select(sample, cell_type), by = c("Sample_name" = "sample")) %>%
  dplyr::select(-c(Sample_name)) %>%
  dplyr::select(Cell_type = cell_type, everything())

heatdata_t_Supp1e <- heatdata_t %>%
  dplyr::select(Sample_name, all_of(selected_genes_Supp1e$Gene[selected_genes_Supp1e$Gene %in% colnames(heatdata_t)])) %>%
  left_join(GRA6_meta %>% dplyr::select(sample, cell_type), by = c("Sample_name" = "sample")) %>%
  dplyr::select(-c(Sample_name)) %>%
  dplyr::select(Cell_type = cell_type, everything())

# Make column annotations
col_annotations <- as_tibble(rownames_to_column(as.data.frame(t(assay(vsd_GRA6))), var = "Sample_name")) %>%
  left_join(GRA6_meta %>% dplyr::select(sample, cell_type), by = c("Sample_name" = "sample")) %>%
  dplyr::select(cell_type) %>%
  pull()
col_annotations <- data.frame(col_annotations)
colnames(col_annotations) <- c("Cell_type")
row.names(col_annotations) <- heatdata_t$Sample_name # Give annotations row names that match column names of heatdata

# Transpose gene expression and rename samples
heatdata_Main1e <- t(heatdata_t_Main1e %>% dplyr::select(-c(Cell_type)))
colnames(heatdata_Main1e) <- heatdata_t$Sample_name

heatdata_Main1g <- t(heatdata_t_Main1g %>% dplyr::select(-c(Cell_type)))
colnames(heatdata_Main1g) <- heatdata_t$Sample_name

heatdata_Supp1e <- t(heatdata_t_Supp1e %>% dplyr::select(-c(Cell_type)))
colnames(heatdata_Supp1e) <- heatdata_t$Sample_name
```

```{r Make row annotations}

row_annotations_Main1e <- data.frame(Annotation = selected_genes_Main1e$Annotation)
row.names(row_annotations_Main1e) <- selected_genes_Main1e$Gene

row_annotations_Main1g <- data.frame(Annotation = selected_genes_Main1g$Annotation)
row.names(row_annotations_Main1g) <- selected_genes_Main1g$Gene

row_annotations_Supp1e <- data.frame(Annotation = selected_genes_Supp1e$Annotation)
row.names(row_annotations_Supp1e) <- selected_genes_Supp1e$Gene
```

```{r Find gap locations}

gaps_Main1e <- selected_genes_Main1e %>%
  dplyr::count(Annotation) %>%
  filter(Annotation == "Transcription Factors") %>%
  select(n) %>%
  pull()

gaps_Main1g <- selected_genes_Main1g %>%
  mutate(Annotation = factor(Annotation, 
                             levels = c("Oxidative Phosphorylation", "Cholesterol Metabolism", "Fatty Acid Metabolism", "Ion Binding", "Transcriptional Regulation"))) %>%
  dplyr::count(Annotation) %>%
  arrange(Annotation) %>%
  mutate(Gaps = cumsum(n)) %>%
  select(Gaps) %>%
  pull()

gaps_Supp1e <- selected_genes_Supp1e %>%
  mutate(Annotation = factor(Annotation,
                             levels = c("CTL Function", "TLRs", "Co-stimulation", "Migration", "Transcription Factors", "Cytokine Receptors"))) %>%
  dplyr::count(Annotation) %>%
  arrange(Annotation) %>%
  mutate(Gaps = cumsum(n)) %>%
  select(Gaps) %>%
  pull()
```

```{r View heatmaps of selected genes}
# Keep columns in the same order as the full de list

# Plot Main1e
Main1e_heatmap <- pheatmap(heatdata_Main1e[,de_order_cols],
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = col_annotations_arranged_de,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(effector = "red", intermediate = "blue", memory = "black")),
         annotation_row = row_annotations_Main1e,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Main1e,
         scale = "row", 
         cellheight = 5, 
         cellwidth = 5, 
         treeheight_col = 10,
         treeheight_row = 0 
  )

# Plot Main1g
Main1g_heatmap <- pheatmap(heatdata_Main1g[,de_order_cols],
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = col_annotations_arranged_de,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(effector = "red", intermediate = "blue", memory = "black")),
         annotation_row = row_annotations_Main1g,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Main1g,
         scale = "row",
         cellheight = 6, 
         cellwidth = 6, 
         treeheight_col = 10,
         treeheight_row = 0 
  )

# Plot Supp1e
Supp1e_heatmap <- pheatmap(heatdata_Supp1e[,de_order_cols],
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = col_annotations_arranged_de,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(effector = "red", intermediate = "blue", memory = "black")),
         annotation_row = row_annotations_Supp1e,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Supp1e,
         scale = "row", 
         cellheight = 4, 
         cellwidth = 6, 
         treeheight_col = 10,
         treeheight_row = 0 
  )
```

```{r Save heatmaps of selected genes}

# Plot Main1e
Main1e_heatmap <- pheatmap(heatdata_Main1e[,de_order_cols], 
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = col_annotations_arranged_de,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(effector = "red", intermediate = "blue", memory = "black")),
         annotation_row = row_annotations_Main1e,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Main1e,
         scale = "row", 
         cellheight = 10,
         cellwidth = 20, 
         filename = "figures/heatmap_selectedgenes_Main1e.png"
  )

# Plot Main1g
Main1g_heatmap <- pheatmap(heatdata_Main1g[,de_order_cols],
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = col_annotations_arranged_de,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(effector = "red", intermediate = "blue", memory = "black")),
         annotation_row = row_annotations_Main1g,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Main1g,
         scale = "row",
         cellheight = 10, 
         cellwidth = 20, 
         filename = "figures/heatmap_selectedgenes_Main1g.png"
  )

# Plot Supp1e
Supp1e_heatmap <- pheatmap(heatdata_Supp1e[,de_order_cols],
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = col_annotations_arranged_de,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(effector = "red", intermediate = "blue", memory = "black")),
         annotation_row = row_annotations_Supp1e,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Supp1e,
         scale = "row", 
         cellheight = 10,
         cellwidth = 20, 
         filename = "figures/heatmap_selectedgenes_Supp1e.png"
  )
```

## Heatmap of selected genes for Hudson et al. data

```{r Load Hudson data}
# Data from: Hudson, W. H. et al. (2019) ‘Proliferating Transitory T Cells with an Effector-like Transcriptional Signature Emerge from PD-1 + Stem- Article Proliferating Transitory T Cells with an Effector-like Transcriptional Signature Emerge from PD-1 + Stem-like CD8 + T Cells during Chronic In’, Immunity. Elsevier Inc., pp. 1–16. doi: 10.1016/j.immuni.2019.11.002.

# Load normalized expression data from Hudson et al.
Hudson_early <- read_excel('/data/yosef2/users/zsteier/Toxo/analysis/GSEA/sig_data/DE Genes Hudson early.xlsx', na = "NA", sheet = 1)
Hudson_late <- read_excel('/data/yosef2/users/zsteier/Toxo/analysis/GSEA/sig_data/DE genes Hudson late.xlsx', na = "NA", sheet = 1)
```

```{r Create Hudson metadata}

# Stem (similar to Tmem): CD101-Tim3-
# Transitory (similary to Tint): CD101-Tim3+
# Exhausted (similar to Teff): CD101+Tim3+

# Create metadata
Hudson_early_meta <- tibble(sample = colnames(Hudson_early)[3:14]) %>%
  mutate(Cell_type = c(rep("Naive", 3), rep("Exhausted", 3), rep("Stem", 3), rep("Transitory", 3))) %>%
  filter(Cell_type != "Naive") %>%
  arrange(factor(Cell_type, levels = c("Stem", "Transitory", "Exhausted")))

Hudson_late_meta <- tibble(sample = colnames(Hudson_late)[3:14]) %>%
  mutate(Cell_type = c(rep("Naive", 3), rep("Stem", 3), rep("Transitory", 3), rep("Exhausted", 3)))%>%
  filter(Cell_type != "Naive") %>%
  arrange(factor(Cell_type, levels = c("Stem", "Transitory", "Exhausted")))

Hudson_early_meta_df <- data.frame(Hudson_early_meta, row.names = "sample")
Hudson_late_meta_df <- data.frame(Hudson_late_meta, row.names = "sample")
```


```{r Select Hudson gene expression data to plot}
# Get expression data 

# Hudson early Main1g
heatdata_Hudsonearly_Main1g <- Hudson_early %>%
  select(c("Gene name", starts_with("CD101"))) %>%
  filter(`Gene name` %in% selected_genes_Main1g$Gene) %>%
  arrange(match(`Gene name`, selected_genes_Main1g$Gene)) %>% 
  select(c("Gene name", row.names(Hudson_early_meta_df))) # arrange columns in correct order

heatdata_Hudsonearly_Main1g_df <- data.frame(heatdata_Hudsonearly_Main1g %>% select(-c("Gene name")), row.names = heatdata_Hudsonearly_Main1g$`Gene name`)
colnames(heatdata_Hudsonearly_Main1g_df) <- colnames(heatdata_Hudsonearly_Main1g %>% select(-c("Gene name")))

# Hudson late Main1g
heatdata_Hudsonlate_Main1g <- Hudson_late %>%
  select(c("Gene name", starts_with("CD101"))) %>%
  filter(`Gene name` %in% selected_genes_Main1g$Gene) %>%
  arrange(match(`Gene name`, selected_genes_Main1g$Gene)) %>% 
  select(c("Gene name", row.names(Hudson_late_meta_df))) # arrange columns in correct order

heatdata_Hudsonlate_Main1g_df <- data.frame(heatdata_Hudsonlate_Main1g %>% select(-c("Gene name")), row.names = heatdata_Hudsonlate_Main1g$`Gene name`)
colnames(heatdata_Hudsonlate_Main1g_df) <- colnames(heatdata_Hudsonlate_Main1g %>% select(-c("Gene name")))

# Hudson early Supp1e
heatdata_Hudsonearly_Supp1e <- Hudson_early %>%
  select(c("Gene name", starts_with("CD101"))) %>%
  filter(`Gene name` %in% selected_genes_Supp1e$Gene) %>%
  arrange(match(`Gene name`, selected_genes_Supp1e$Gene)) %>% 
  select(c("Gene name", row.names(Hudson_early_meta_df))) # arrange columns in correct order

heatdata_Hudsonearly_Supp1e_df <- data.frame(heatdata_Hudsonearly_Supp1e %>% select(-c("Gene name")), row.names = heatdata_Hudsonearly_Supp1e$`Gene name`)
colnames(heatdata_Hudsonearly_Supp1e_df) <- colnames(heatdata_Hudsonearly_Supp1e %>% select(-c("Gene name")))

# Hudson late Supp1e
heatdata_Hudsonlate_Supp1e <- Hudson_late %>%
  select(c("Gene name", starts_with("CD101"))) %>%
  filter(`Gene name` %in% selected_genes_Supp1e$Gene) %>%
  arrange(match(`Gene name`, selected_genes_Supp1e$Gene)) %>% 
  select(c("Gene name", row.names(Hudson_late_meta_df))) # arrange columns in correct order

heatdata_Hudsonlate_Supp1e_df <- data.frame(heatdata_Hudsonlate_Supp1e %>% select(-c("Gene name")), row.names = heatdata_Hudsonlate_Supp1e$`Gene name`)
colnames(heatdata_Hudsonlate_Supp1e_df) <- colnames(heatdata_Hudsonlate_Supp1e %>% select(-c("Gene name")))
```

```{r View Hudson heatmaps}

# Plot Main1e
Main1g_heatmap_Hudsonearly <- pheatmap(heatdata_Hudsonearly_Main1g_df,
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = Hudson_early_meta_df,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(Exhausted = "red", Transitory = "blue", Stem = "black")),
         annotation_row = row_annotations_Main1g,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Main1g,
         scale = "row", 
         cellheight = 6,
         cellwidth = 6, 
         treeheight_col = 10,
         treeheight_row = 0 
  )

# Plot Main1e
Main1g_heatmap_Hudsonlate <- pheatmap(heatdata_Hudsonlate_Main1g_df,
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = Hudson_late_meta_df,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(Exhausted = "red", Transitory = "blue", Stem = "black")),
         annotation_row = row_annotations_Main1g,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Main1g,
         scale = "row", 
         cellheight = 6, 
         cellwidth = 6, 
         treeheight_col = 10,
         treeheight_row = 0 
  )

# Plot Supp1e
Supp1e_heatmap_Hudsonearly <- pheatmap(heatdata_Hudsonearly_Supp1e_df,
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = Hudson_early_meta_df,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(Exhausted = "red", Transitory = "blue", Stem = "black")),
         annotation_row = row_annotations_Supp1e,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Supp1e,
         scale = "row", 
         cellheight = 4, 
         cellwidth = 6, 
         treeheight_col = 10,
         treeheight_row = 0 
  )

# Plot Supp1e
Supp1e_heatmap_Hudsonlate <- pheatmap(heatdata_Hudsonlate_Supp1e_df,
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = Hudson_late_meta_df,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(Exhausted = "red", Transitory = "blue", Stem = "black")),
         annotation_row = row_annotations_Supp1e,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Supp1e,
         scale = "row", 
         cellheight = 4, 
         cellwidth = 6, 
         treeheight_col = 10,
         treeheight_row = 0 
  )
```

```{r Save Hudson heatmaps}

# Plot Main1g
Main1g_heatmap <- pheatmap(heatdata_Hudsonlate_Main1g_df,
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = Hudson_late_meta_df,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(Exhausted = "red", Transitory = "blue", Stem = "black")),
         annotation_row = row_annotations_Main1g,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Main1g,
         scale = "row",
         cellheight = 10, 
         cellwidth = 20, 
         filename = "figures/heatmap_selectedgenes_Hudsonlate_Main1g.png"
  )

# Plot Supp1e
Supp1e_heatmap <- pheatmap(heatdata_Hudsonlate_Supp1e_df,
         cluster_rows = FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = Hudson_late_meta_df,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(Exhausted = "red", Transitory = "blue", Stem = "black")),
         annotation_row = row_annotations_Supp1e,
         border_color = NA, 
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100), 
         gaps_row = gaps_Supp1e,
         scale = "row", 
         cellheight = 10,
         cellwidth = 20, 
         filename = "figures/heatmap_selectedgenes_Hudsonlate_Supp1e.png"
  )
```

## Heatmap of selected DE genes for Hudson et al. data

```{r Select DE genes to plot}

# Filter for genes upregulated specifically in Tint (DE in both pairwise tests)
de_int <- GRA6_DE %>%
  filter( ((memory_vs_intermediate_padj < .05) & (memory_vs_intermediate_log2FoldChange < 1)) & 
            ((memory_vs_effector_padj < .05) & (intermediate_vs_effector_log2FoldChange > 1)) ) %>%
  dplyr::select(Gene) %>%
  pull()

length(de_int)
```

```{r Prepare Hudson expression data for Tint genes}

heatdata_Hudsonlate_Tint <- Hudson_late %>%
  select(c("Gene name", starts_with("CD101"))) %>%
  filter(`Gene name` %in% de_int) %>%
  rowwise() %>%
  mutate(counts = sum(c(`CD101-Tim3-...6`:`CD101+Tim3+...14`))) %>% # Filter out genes with zero counts
  select(c(`Gene name`, counts, everything())) %>%
  filter(counts > 0) %>%
  select(-c("counts")) %>%
  arrange(match(`Gene name`, de_int)) %>% 
  select(c("Gene name", row.names(Hudson_late_meta_df))) # arrange columns in correct order

heatdata_Hudsonlate_Tint_df <- data.frame(heatdata_Hudsonlate_Tint %>% select(-c("Gene name")), row.names = heatdata_Hudsonlate_Tint$`Gene name`)
colnames(heatdata_Hudsonlate_Tint_df) <- colnames(heatdata_Hudsonlate_Tint %>% select(-c("Gene name")))
```
```{r Plot Hudson heatmap for Tint genes}

# Save
pheatmap(heatdata_Hudsonlate_Tint_df,
         cluster_rows = TRUE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = Hudson_late_meta_df,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(Exhausted = "red", Transitory = "blue", Stem = "black")),
         annotation_names_col = TRUE,
         border_color = NA, 
         treeheight_row = 0,
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100),
         scale = "row", 
         cellheight = 10, 
         cellwidth = 20, 
         filename = str_c("figures/heatmap_Tint_Hudsonlate_Supp1d.png")
  )

# View
pheatmap(heatdata_Hudsonlate_Tint_df,
         cluster_rows = TRUE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         cluster_cols = FALSE,
         annotation_col = Hudson_late_meta_df,
         angle_col = 90,
         annotation_colors = list("Cell_type" = c(Exhausted = "red", Transitory = "blue", Stem = "black")),
         annotation_names_col = TRUE,
         border_color = NA, 
         color = colorRampPalette(rev(brewer.pal(10, "RdBu")))(100),
         scale = "row", 
         cellheight = 4, 
         cellwidth = 15
  )
```

## GSEA results as triwise plots

```{r Load custom signatures to plot}

# Hudson_late_CD101negTim3neg_vs_CD101negTim3pos (Stem signature)
late_CD101negTim3neg_vs_CD101negTim3pos <- read_excel('/data/yosef2/users/zsteier/Toxo/analysis/GSEA/sig_data/DE genes Hudson late.xlsx', na = "NA", sheet = 3)

# Make sig direction tibble (Gene_Symbol, Direction)
Hudson_late_CD101negTim3neg_vs_CD101negTim3pos <- late_CD101negTim3neg_vs_CD101negTim3pos %>%
  filter(abs(log2FoldChange) > 1) %>% # filter on LFC
  filter(padj < 0.05) %>% # filter on padj
  mutate(Gene_Symbol = external_gene_name) %>%
  mutate(Direction = case_when(log2FoldChange > 0 ~ "Up",
                               log2FoldChange < 0 ~ "Down")) # 2,493 genes
```

```{r Collect gene expression data}

# Collect gene expression data (VST normalized)
vsd_GRA6_all <- vst(dedsRNASeq_GRA6, blind=FALSE)
heatdata_t <- as_tibble(rownames_to_column(as.data.frame(t(assay(vsd_GRA6_all))), var = "Sample_name")) 

# Add metadata
heatdata_tnew <- heatdata_t %>%
  left_join(GRA6_meta %>% dplyr::select(sample, cell_type), by = c("Sample_name" = "sample")) %>%
  dplyr::select(-c(Sample_name)) %>%
  dplyr::select(Cell_type = cell_type, everything())

# Average the expression per condition
summarized_gra6 <- heatdata_tnew %>%
  group_by(Cell_type) %>%
  summarize(across(everything(), mean)) 

# transform to gene x cell type matrix
summarized_gra6_t <- as.matrix(t(summarized_gra6 %>% dplyr::select(-c("Cell_type"))))
colnames(summarized_gra6_t) <- c("effector", "intermediate", "memory")
```

```{r Collect genes in stem signature for triwise plot}

# Hudson_late_CD101negTim3neg_vs_CD101negTim3pos (Stem signature: Stem vs transitory)
Stem_sig_up <- Hudson_late_CD101negTim3neg_vs_CD101negTim3pos %>%
  dplyr::select(Gene_Symbol, Direction) %>%
  filter(Direction == "Up", Gene_Symbol %in% row.names(summarized_gra6_t)) %>%
  dplyr::select(Gene_Symbol) %>%
  pull()
Stem_sig_down <- Hudson_late_CD101negTim3neg_vs_CD101negTim3pos %>%
  dplyr::select(Gene_Symbol, Direction) %>%
  filter(Direction == "Down", Gene_Symbol %in% row.names(summarized_gra6_t)) %>%
  dplyr::select(Gene_Symbol) %>%
  pull()
Stem_sig_all <- unique(c(Stem_sig_up, Stem_sig_down))
```

```{r Plot Hudson Stem sig as triwise}

gene_set_expression <- summarized_gra6_t[Stem_sig_up,]
dim(gene_set_expression)

# transform into barycentric coordinates (from genes x sample)
barycoords = transformBarycentric(gene_set_expression)

# plot as hexagonal plot (triwise plot)
plotDotplot(barycoords, Gdiffexp = rownames(barycoords), 
            baseangle = -.5,
            colorvalues = c("black"),
            rmax = 4, showlabels = FALSE)
ggsave("figures/triwise_Hudson_late_Stem_up.png", dpi = 600, bg = "transparent")
```